---
title: "Data Wrangling with R: Day 1"
subtitle: "Data manipulation with `dplyr`"
author: "Emi Tanaka"
email: "emi.tanaka@monash.edu"
date: "1st December 2020 @ Statistical Society of Australia | Zoom"
length: "30 minutes"
titleslide_bgimg: "images/bg2.jpg"
exercise: "day1-exercise-02"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/animate.css"
      - "assets/custom.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
    self_contained: false 
    seal: false 
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    mathjax: "assets/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: github
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
library(flair)
current_file <- knitr::current_input()
flair_on <- TRUE
```
```{r titleslide, child="assets/titleslide.Rmd"}
```

---

# Grammar of data manipulation

```{r, warning = FALSE, message = FALSE}
library(dplyr) # or library(tidyverse)
```
* `dplyr` is a core package in `tidyverse` 
* The earlier concept of `dplyr` (first on CRAN in 2014-01-29) was implemented in `plyr` (first on CRAN in 2008-10-08)
* The functions in `dplyr` has been evolving frequently but `dplyr` v1.0.0 was released on CRAN in 2020-05-29
* This new version contained new "verbs"
* The major release suggests that functions in `dplyr` are maturing and thus the user interface is unlikely to change





.footnote[
Hadley Wickham, Romain François, Lionel Henry and Kirill Müller
  (2020). dplyr: A Grammar of Data Manipulation. R package version
  1.0.2. 
  
Hadley Wickham (2011). The Split-Apply-Combine Strategy for Data
  Analysis. Journal of Statistical Software, 40(1), 1-29.
]


---

# Lifecycle

.grid[
.item[
<br>
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/vignettes/figures/lifecycle.svg" width = "550px">
]
.item[

<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-archived.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-defunct.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-deprecated.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-experimental.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-maturing.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-questioning.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-retired.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-soft-deprecated.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-stable.svg">
<img src="https://raw.githubusercontent.com/r-lib/lifecycle/master/man/figures/lifecycle-superseded.svg">

* Functions (and sometimes arguments of functions) in `tidyverse` packages often are labelled with a badge like above
* Find definitions of badges [here](https://lifecycle.r-lib.org/articles/lifecycle.html)
* Check out documentations below
```{r, eval = FALSE}
help(mutate, package = "dplyr")
help(mutate_each, package = "dplyr")
```

]

]

.footnote[
Lionel Henry (2020). lifecycle: Manage the Life Cycle of your
  Package Functions. R package version 0.2.0.
]


---

# `dplyr` "verbs"


* The main functions of `dplyr` include: 
<center>
<table style="width:60%">
  <tr>
    <td><code>arrange</code></td>
    <td><code>select</code></td>
    <td><code>mutate</code></td>
  </tr>
  <tr>
    <td><code>rename</code></td>
    <td><code>group_by</code></td>
    <td><code>summarise</code></td>
  </tr>
</table>
</center>
--

* Notice that these functions are _verbs_
--

* Functions in `dplyr` generally have the form:

.center[
`verb(data, args)`
]

--

* I.e., the first argument `data` is a `data.frame` object
--

* What do you think the following will do?

.center[
`rename(mtcars, miles_per_gallon = mpg)`

{{content}}
]
--

`arrange(mtcars, wt)`

---

# Pipe operator %>%

* Almost all tidyverse packages import the `magrittr` package to use `%>%`

.footnote[
Stefan Milton Bache and Hadley Wickham (2020). magrittr: A
  Forward-Pipe Operator for R. R package version 2.0.1.
]
--

* `x %>% f(y)` is the same as `f(x, y)`
--

* `x %>% f(y) %>% g(z)` is the same as `g(f(x, y), z)`
--

* When you see the pipe operator `%>%`, read it as "and then"
--

```{r}
mtcars %>%                          # take mtcars data, and then
 rename(miles_per_gallon = mpg) %>% # rename mpg as miles_per_gallon, and then
 arrange(wt)                        # arrange row by wt
```

---

# Lazy and non-standard evaluation

* Remember in Base R:

```{r}
subset(mtcars, mpg > 31)
```
--

* But the second argument cannot be evaluated:

```{r, error = TRUE}
mpg > 31
```
--

* R employs what is called **lazy evaluation** for function inputs 
--

* **Non-standard evaluation** uses this feature to capture the input expression within the function and evaluate only when requested

---

# Tidy evaluation .font_small[.font_small[Part] 1]

* Tidy evaluation builds on the lazy and non-standard evaluation and is implemented in `rlang`
* All core tidyverse packages import `rlang`
* So what does it do?
* Let's consider `filter`, the Tidyverse version of `subset`

```{r}
filter(mtcars, mpg > 31)
```

* <i class="fas fa-exclamation-triangle animated flash red"></i> If you get an error using `filter`, replace it with `dplyr::filter`<br> .font_small[for those interested, `dplyr::filter` is a conflict with `stats::filter` and it may be using `stats::filter` instead... I've fallen into this trap so many times!]

.footnote[
Lionel Henry and Hadley Wickham (2020). rlang: Functions for Base
  Types and Core R and 'Tidyverse' Features. R package version
  0.4.8.
]

---

# Tidy evaluation .font_small[.font_small[Part] 2]

* Suppose we have a silly function that subsets `mtcars` for a given condition

```{r}
myCarSubset <- function(cond) subset(mtcars, cond)
myCarFilter <- function(cond) filter(mtcars, cond)
```

* This causes an issue because `cond` is evaluated before it is parsed into `subset` or `filter`

```{r, error = TRUE}
myCarSubset(mpg > 31)
myCarFilter(mpg > 31)
```

---

# Tidy evaluation .font_small[.font_small[Part] 3]

* Functions that use non-standard evaluation is problematic

```{r curly, error = TRUE, eval = !flair_on, echo = !flair_on}
myCarSubsetNew <- function(cond) subset(mtcars, {{ cond }})
myCarFilterNew <- function(cond) filter(mtcars, {{ cond }})

myCarSubsetNew(mpg > 31)
myCarFilterNew(mpg > 31)
```
```{r, error = TRUE, echo = FALSE, eval = flair_on}
decorate("curly") %>% 
  flair("{{") %>% 
  flair("}}")
```


* .monash-blue[`{{ }}`] only works if the underlying function implements `rlang`

---

# Data masking .font_small[.font_small[Part] 1]

.grid[
.item[
```{r, error = TRUE}
ind <- 1:nrow(cars) # nrow(cars) = 50
subset(cars, ind > 49)
filter(cars, ind > 49)
```
* For any variables that don't exist in the data, R searches the parental environment for evaluation.
]
.item[

{{content}}
]

]


--
```{r subset, eval = FALSE}
speed <- c(40, 51)
subset(cars, speed > 24)
```
{{content}}
--
```{r subset, echo = FALSE}
```
{{content}}
--
```{r filter, eval = FALSE}
filter(cars, speed > 24)
```
{{content}}
--
```{r filter, echo = FALSE}
```
{{content}}
--

* The variables in data take priority for reference over those in parental environment

---

# Data masking .font_small[.font_small[Part] 2]


.grid[
.item[
```{r data-mask, eval = !flair_on, echo = !flair_on}
speed <- 1:nrow(cars)
filter(cars, .data$speed > 24)
filter(cars, .env$speed > 24)
```

```{r, echo = FALSE, eval = flair_on}
decorate("data-mask") %>% 
  flair(".data") %>% 
  flair(".env")
```

]
.item[

* In Tidyverse, you can be explicit whether the variable is in the data or in the parental environment
* .monash-blue[`.data`] is a special pronoun referring to variables in the data parsed in the first argument
* .monash-blue[`.env`] is a special pronoun referring to variables in the environment (i.e. _not_ in the data parsed in the first argument)

]

]

---

# Tidy select .font_small[.font_small[Part] 1]

* Tidyverse packages generally use syntax from the `tidyselect` package for variable/column selection

```{r}
data(frog_signal, package = "dwexercise")
str(frog_signal) 
```

---

# Tidy select .font_small[.font_small[Part] 2]

.grid[
.item50[
```{r select, eval = !flair_on, echo = !flair_on}
frog_signal %>% 
 select(Standard1, Standard2, Standard3)
```
```{r, echo = FALSE, eval = flair_on}
decorate("select") %>% 
  flair("select") 
```

]
.item50[
{{content}}
]

]
--

```{r select2, eval = !flair_on, echo = !flair_on}
frog_signal %>% 
  select(Standard1:Standard3)
```
```{r, echo = FALSE, eval = flair_on}
decorate("select2") %>% 
  flair(":") 
```

<div class="info-box pad20" style="position:absolute;right:10px;bottom:10px;">
The <code>tidyselect</code> syntax <code class="monash-blue">:</code> can be used to select contiguous columns in the data
</div>

---

# Tidy select .font_small[.font_small[Part] 3]

.grid[
.item50[
```{r}
frog_signal %>% 
  select(starts_with("Standard"))
```
]
.item[
{{content}}
]

]

--

```{r}
frog_signal %>% 
  select(num_range("Standard", 1:3))
```

---

# Selection language .font_small[.font_small[Part] 1]

* `:` for selecting contiguous variables
* `!` for taking complement set of variables
* `&` or `|` for selecting intersection or union of two sets of variables, e.g. 
```{r}
frog_signal %>% 
  select(starts_with("Alt") & ends_with("1")) %>% 
  str()
```

* `c()` for combining selections
* `everything()` to select all variables
* `last_col()` to select last variable, with option of an offset

---

# Selection language .font_small[.font_small[Part] 2]


* `starts_with()` selects columns with the given prefix
* `ends_with()` selects columns with the given suffix
* `contains()` selects columns with a literal string
* `matches()` selects columns that match the regular expression .font_small[we'll learn this next!]
* `num_range()` selects columns with a numerical range
* `all_of()` for selecting columns based on a character vector
* `any_of()` is the same as `all_of()` but no error when variables do not exist
* `where()` selects based on where given function return TRUE

```{r, eval = FALSE}
help(language, package = "tidyselect")
```


---

# Subsetting by column .font_small[Tidyverse]

.grid[
.item[
```{r, eval = FALSE}
select(mtcars, c(mpg, cyl))
select(mtcars, c("mpg", "cyl"))
select(mtcars, mpg, cyl)
select(mtcars, "mpg", "cyl")
```
All the same result as below
```{r}
mtcars %>% 
  select(mpg, cyl)
```

]
.item[
{{content}}
]

]

--

```{r}
mtcars %>% select(mpg)
```

--

<div class="info-box" style="position:absolute;right:20px;margin-right:0px!important;bottom:0px;margin-left:0;width:650px;">
<ul>
<li>Selecting one column doesn't "drop" it to a vector.</li> <li>If you really want the vector then use <code class="monash-blue">pull(mpg)</code>.</li>
</ul>
</div>


---

# Subsetting by row .font_small[Tidyverse]

```{r}
mtcars %>% 
  slice(3:1)
```
--
```{r}
mtcars %>% 
  filter(rownames(.) %in% c("Datsun 710", "Mazda RX4"))
```
* Use .monash-blue[`slice`] to subset by *index* and .monash-blue[`filter`] by *logical vector*

--

<div class="info-box" style="position:absolute;right:20px;margin-right:0px!important;top:50px;margin-left:0;width:900px;font-size: 20pt;">
<ul>
<li>Placeholder binding</li>
<ul>
<li><code>x %>% f(y, g(.))</code> is the same as <code>f(x, y, g(x))</code></li>
<li><code>x %>% f(y, .)</code> is the same as <code>f(y, x)</code></li>
</ul>
{{content}}
</ul>
</div>

--
<i class="fas fa-exclamation-triangle red"></i> Note: row names do not follow tidy data principles<br>
{{content}}
--
<li>Use <code class="monash-blue">tibble::rownames_to_column()</code> to convert rownames to a column to make into a tidy data</li>



---

# Adding or modifying a column .font_small[Tidyverse]

```{r mutate, eval = !flair_on, echo = !flair_on}
mtcars %>% 
  mutate(gpm = 1 / mpg,
          wt = gpm^2,
          wt = if_else(cyl==6, 10, wt),
          hp = case_when(cyl==6 ~ 11,
                         cyl==4 ~ 10,
                           TRUE ~ 3))
```
```{r, echo = FALSE, eval = flair_on}
decorate("mutate") %>% 
  flair("mutate")
```
--

<div class="info-box" style="position:absolute;right:20px;margin-right:0px!important;top:50px;margin-left:0;width:550px;font-size: 20pt;">
<ul>
<li>Evaluation in <code>mutate</code> is done sequentially based on input order</li>
{{content}}
</ul>
</div>

--

<li>So you refer to the newly created variable in later input</li>
{{content}}
--
<li>You can call multiple <code>mutate</code> but computational performance is usually better if done within the same <code>mutate</code> call</li>
<pre>
<code>mtcars %>%
  mutate(gpm = 1 / mpg) %>% 
  mutate(wt = gpm^2)
</code></pre>

---

# Sorting columns .font_small[Tidyverse]

.grid[
.item50[
.s300[
```{r}
mtcars %>% 
  select(sort(names(.)))
```
]
.s300[
```{r}
mtcars %>% 
  relocate(am, carb, .before = cyl)
```
]
]
.item50[
.s300[
```{r}
mtcars %>% 
  select(wt, gear, everything())
```
]
.s300[
```{r}
mtcars %>% 
  relocate(wt, gear, .after = mpg)
```
]
]

]

---

# Calculating statistical summaries by group .font_small[Tidyverse]

--

.grid[.item50[

🎯 Calculate the _average_ weight (`wt`) of a car for each gear type in (`gear`)  `mtcars`
```{r}
mtcars %>% 
  group_by(gear) %>% 
  summarise(avg_wt = mean(wt))
```

]
.item[
{{content}}
]
]
--

🎯 Calculate the _median_ weight (`wt`) of a car for each gear (`gear`) and engine (`vs`) type in  `mtcars`

```{r}
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(avg_wt = mean(wt),
            med_wt = median(wt))
```

---

#  `across` .tag.flash.animated[NEW in `dplyr` v1.0.0]

* Using `across`, you can more easily apply a function to multiple columns

```{r, message = FALSE}
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(across(everything(), mean)) #<<
```

---

#  `where` .tag.flash.animated[NEW in `dplyr` v1.0.0]

* You can combine `across` with the selection helper `where` 

```{r, message = FALSE}
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(across(where(function(x) n_distinct(x) > 10), mean)) #<<
```

---

#  `c_across` .tag.flash.animated[NEW in `dplyr` v1.0.0]

* Remember tidy selection only works with functions that are compatible

```{r, message = FALSE}
mtcars %>% 
  rowwise() %>% 
  summarise(disp = disp, hp = hp, drat = drat, wt = wt,
            score = sum(c_across(disp:wt))) #<<
```
---

#  `rowwise`?

* What happens if you omit `rowwise`?

```{r, message = FALSE}
mtcars %>% 
  #rowwise() %>% 
  summarise(disp = disp, hp = hp, drat = drat, wt = wt,
            score = sum(c_across(disp:wt))) #<<
```

---

```{r exerciseslide, child="assets/exerciseslide.Rmd"}
```

---

```{r endslide, child="assets/endslide.Rmd"}
```
