---
title: "Data Wrangling with R: Day 1"
author: "Emi Tanaka"
date: "1st December 2020"
---

## Slides

We recommend that you open the html slides <i class="fas fa-link"></i> with [Google Chrome web browser](https://www.google.com/intl/en_au/chrome/). 

0. Getting started [<i class="fas fa-link"></i>](slides/day1-session0.html) [<i class="fas fa-file-pdf"></i>](slides/day1-session0.pdf)
1. Base R and `tidyverse` [<i class="fas fa-link"></i>](slides/day1-session1.html) [<i class="fas fa-file-pdf"></i>](slides/day1-session1.pdf)
2. Data manipulation with `dplyr` [<i class="fas fa-link"></i>](slides/day1-session2.html) [<i class="fas fa-file-pdf"></i>](slides/day1-session2.pdf)
3. Manipulating strings with `stringr` [<i class="fas fa-link"></i>](slides/day1-session3.html) [<i class="fas fa-file-pdf"></i>](slides/day1-session3.pdf)
4. Tidying data with `tidyr` [<i class="fas fa-link"></i>](slides/day1-session4.html) [<i class="fas fa-file-pdf"></i>](slides/day1-session4.pdf)

## Schedule

```{r day1, echo = FALSE, results="asis"}
cat(readLines("schedule/schedule-day1.md"), sep = "\n")
```

## Exercises

Please make sure you install the following packages with latest versions.

```` r
```{r pkgs, echo = FALSE, results="asis", warning = FALSE}
pkgs <- readLines("install-pkgs.txt")
rmts <- readLines("install-remotes.txt")
cat(glue::glue("install.packages(\"{pkgs}\")"), sep = "\n")
cat("\n# you will need to have installed the remotes package for below\n")
cat(glue::glue("remotes::install_github(\"{rmts}\")"), sep = "\n")

```
````

To do the exercises, please install the following package:

```r
remotes::install_github("emitanaka/dwexercise")
```

In order to launch the exercises, use one of the commands below:

```r
learnr::run_tutorial("day1-exercise-01", package = "dwexercise")
learnr::run_tutorial("day1-exercise-02", package = "dwexercise")
learnr::run_tutorial("day1-exercise-03", package = "dwexercise")
learnr::run_tutorial("day1-exercise-04", package = "dwexercise")
```

If the above are not working for you, you can try the shiny app version instead:

* https://ebsmonash.shinyapps.io/dw-day1-exercise-01
* https://ebsmonash.shinyapps.io/dw-day1-exercise-02
* https://ebsmonash.shinyapps.io/dw-day1-exercise-03
* https://ebsmonash.shinyapps.io/dw-day1-exercise-04
